A fast GPU solver for near-Toeplitz tridiagonal systems

### 2015-11-07 paper-draft/paper.tex
fix formatting issues

\begin{abstract}
    In this paper, we consider the solution of linear
    systems of the form $A\bm{x} = \bm{d}$,
    where $A$ is \emph{near-Toeplitz}.
    These systems appear in numerical schemes
    such as
    Alternating Direction Implicit (ADI) methods,
    compact finite differences,
    and one-dimensional ordinary
    and partial differential equations.
    We develop a GPU solver that takes advantage
    of the known structure of the matrix
    to reduce the number of computations
    and improve memory access.
    Our solver is able to achieve up to 5x speedup
    over an Intel MKL solver running on a single core,
    and 1.5x-11x speedup
    over the NVIDIA CUSPARSE solver.
    For larger problem sizes,
    our solver is able to outperform
    CUSPARSE as well as Intel MKL (on up to 16 cores).
    We also describe the implementation of a multi-GPU solver
    using MPI and CUDA.
    We describe two applications:
    2-D ADI applied to the heat equation on a single GPU,
    and computation of
    3-D compact finite differences on multiple GPUs.

\end{abstract}

@@

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}

Where, in general
\begin{align*}
    & a_0 \neq a_n & \\
    & b_1 \neq b_0 \neq b_n &\\
    & c_1 \neq c_0 &
\end{align*}

@@

Such tridiagonal systems
appear in several numerical schemes
in Computational Fluid Dynamics (CFD)
such as
compact finite differences \cite{lele1992compact},
Alternating Direction Implicit (ADI) methods \cite{1955ADI}, and
numerical solutions of one-dimensional differential equations.
In such applications,
the evaluation of the tridiagonal systems constitutes
a significant portion of the runtime.
Recently, there has been considerable interest in the CFD community
in developing general tridiagonal solvers that
exploit the massive parallelism afforded by
Graphics Processing Units (GPUs)
to speed up the computations
\cite{tutkun2012gpu}
\cite{esfahanian2014efficient}
\cite{GoSt11CR}.
The performance of tridiagonal solvers on the GPU
has been shown to depend on several factors,
such as
choice of algorithm and the associated complexity,
thread utilization,
use of the available memory hierarchy,
and synchronization and control costs
\cite{Zhang2010FTS}.
Many approaches described in the literature
seek to make trade-offs between these factors
to arrive at a solver that best fits the requirements.
But to our knowledge,
the algorithms proposed so far
are applicable to general tridiagonal systems,
and make no attempt to take advantage of
the known structure of the tridiagonal matrix.
We describe the implementation of a GPU solver that
uses this information to
reduce the number of computations
and memory accesses.
The resulting solver is thus able to perform twice as fast as
the equivalent NVIDIA CUSPARSE solver,
and an Intel MKL solver running with 16 independent cores.

This paper is organized as follows:
Section \ref{sec:preliminaries}
describes algorithms for solving tridiagonal systems on GPUs.
The cyclic reduction algorithm in particular, is discussed in detail,
and issues with its implementation on GPUs is discussed.
Section \ref{sec:proposed-algorithm}
describes our proposed solver for near-Toeplitz tridiagonal systems.
Section \ref{sec:gpu-implementation}
discusses the implementation of the algorithm on GPUs.
Section \ref{sec:results-single-gpu}
provides an overview of the performance of our solver,
and compares it with other implementations.

@@

Cyclic reduction is applicable to the solution of
general linear systems of the form $Ax = d$,
where $A$ is a tridiagonal matrix with diagonals
$a$, $b$ and $c$
(Equation \ref{eqn:general-tridiagonal-system}).
The algorithm consists of two phases:
\emph{forward reduction} and \emph{backward substitution}.
In the forward reduction phase,
every even-indexed equation $i$
is expressed as a
linear combination of equations $i$, $i-1$ and $i+1$,
yielding a new tridiagonal system of
$n/2$ equations in $n/2$ unknowns
(Equations \ref{eqn:forward-reduction-1} - \ref{eqn:forward-reduction-4}).
The process is repeated until a system of
2 equations in 2 unknowns is left.

\begin{align} 
& a^{\prime}_i = -a_{i-1}k_1 \
    \label{eqn:forward-reduction-1}& \\
& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 \
    \label{eqn:forward-reduction-2}& \\
& c^{\prime}_i = -c_{i+1}k_2 \
    \label{eqn:forward-reduction-3}& \\
& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
    \label{eqn:forward-reduction-4}&
\end{align}

where,

@@

\begin{align} \label{eqn:backward-substitution}
x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
    c^{\prime}_ix_{i+1}}{b^{\prime}_i}
\end{align}

For the last index $i=n$,
the forward reduction step is instead:
\begin{align} \label{eqn:forward-reduction-last}
    & a^{\prime}_n = -a_{n-1}k_1 & \\
    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
    & d^{\prime}_n = d_n - d_{n-1}k_1&
\end{align}

And for $i=1$, the backward substitution step is instead:
\begin{align} \label{eqn:backward-substitution-first}
x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
\end{align}

@@

Writing applications that effectively exploit the GPU
requires a thorough understanding of the underlying architecture.
We provide the pertinent details here,
and refer to \cite{GPUcomputingera} for a more complete picture.
Our tests are performed on the NVIDIA Tesla K20 and K40 GPUs,
built on the ``Kepler'' architecture,
an overview of which is available at \cite{Keplerwhitepaper}.

@@

\begin{itemize}
    \item [] $a_2 = a_3 = a_4 = \hdots \equiv a_0$
    \item [] $b_2 = b_3 = b_4 = \hdots \equiv b_0$
    \item [] $c_2 = c_3 = c_4 = \hdots \equiv c_0$
\end{itemize}

we observe that the resulting coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
correspond to the coefficients of a tridiagonal matrix with
exactly the near-Toeplitz structure of $A$.
This form-preserving property of cyclic reduction
has been reported for block Toeplitz tridiagonal systems
by Bini et al.

@@

\begin{equation}
d^{\prime}_i = d_i - d_{i-1}k_1^{m}  - d_{i+1}k_2^{m}
\label{eqn:precomputed-forward-reduction-step}
\end{equation}

And the $m^{th}$ backward substitution step is:

\begin{equation}
x_i = \frac{d^{\prime}_i - a^mx_{i-1} - \
    c^{m}x_{i+1}}{b^m}
\label{eqn:precomputed-backward-substitution-step}
\end{equation}

Where $a^m$, $b^m$, $c^m$, $k_1^m$ and $k_2^m$
are the appropriate values
from the precomputed coefficient arrays.

@@

Following the algorithm described in \ref{sec:proposed-algorithm},
the forward reduction coefficients
are first computed and loaded into global memory.
The right hand sides are stored in a single
contiguous region global memory.
In both approaches,
we map each of the $N_{rhs}$ tridiagonal systems
to a thread block,
and equations or \emph{indices} of a system
to individual threads within a block.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=400pt]{img/global-and-shared.eps}
\end{center}
\caption{Shared memory (top) and Global memory (bottom) implementations---
    multiple kernel launches allow us to dedicate appropriate number
    of threads per block in the global memory approach}
\label{fig:global-and-shared}
\end{figure}

@@

This implementation works entirely on the GPU's global memory.
Each right-hand side is mapped to a single thread block.
Each block executes two kernels,
\texttt{forwardReduction} and \texttt{backSubstitution},
which are called each called $log_2(n)-1$ times.
An extra call to the
forward reduction kernel additionally performs
the two-by-two solve.
Exit from a kernel marks a global synchronization point,
so no explicit synchronization is performed in the kernels.
Because we launch a kernel for each
forward reduction and back substitution step,
we can specify the appropriate number of threads per block---$n/2$
threads per block for the first forward reduction step,
$n/4$ for the second, and so on (Figure \ref{fig:global-and-shared}).
This guarantees that
all threads within a block are always active.
The precomputed coefficient arrays and right hand
are accessed by the kernels from global memory directly.
Because the coefficients
$k_{1,i}$ and $k_{2,i}$
required in the forward reduction step
(Equation \ref{eqn:precomputed-forward-reduction-step})
are precomputed and available in global memory,
we do not compute
$a_i^\prime$, $b_i^\prime$, $c_i^\prime$,
$k_{1,i}$ and $k_{2,i}$.
This reduces the number of uncoalesced accesses to global memory.
In the backward substitution steps
(Equation \ref{eqn:precomputed-backward-substitution-step}),
the coefficients $a^m$, $b^m$ and $c^m$
are accessed from global memory without coalescing problems.

@@

In the shared memory approach (Figure \ref{fig:global-and-shared}),
we again map every right-hand side to a thread block.
Each thread block executes a single kernel \emph{once},
which performs forward reduction,
backward substitution, and the two-by-two solve.
Shared memory sized $n/2$ is allocated by each block.
Each thread performs the first reduction step
(Equation \ref{eqn:precomputed-forward-reduction-step})
by accessing the required values
$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
storing the result in shared memory.
In subsequent reduction steps,
$d_i$, $d_{i-1}$ and $d_{i+1}$
are accessed from shared memory,
avoiding uncoalesced global memory accesses.
In each back substitution step,
threads overwrite the existing values in shared memory
with the values of the solution.
In the final step,
shared memory is filled completeley
with the even-indexed solution values.
Each thread then computes an odd-indexed solution value,
storing it directly in global memory
and copies the even-indexed solution value
from shared memory to global memory.
The entire solution is done in a single kernel launch
to hide the latency of
memory transfers between global and shared memory.
Explicit synchronization between the threads of a block
is required within the kernel at the end of each step.

@@

\begin{figure}[!htb]
        \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
        \caption{Comparison of global memory and shared
            memory implementations of NEATO (2D problems)}
        \label{fig:global-vs-shared-2d}
        \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
        \caption{Comparison of global memory and shared memory
            implementations of NEATO (2D problems)}
        \label{fig:global-vs-shared-3d}
\end{figure}

@@

In Figures \ref{fig:bench-2d} and \ref{fig:bench-3d},
we provide the relative performance of
Intel MKL and CUSPARSE solvers and compare against
the NEATO shared memory implementation.
The Intel MKL solver can use multiple threads
to solve several systems simaltaneously.
The CUSPARSE solver uses a hybrid algorithm composed
of Cyclic Reduction and Parallel Cyclic Reduction (CR+PCR).
The relative performance for
each problem size is obtained by
normalizing the solver timings
by the timing for the fastest solver for that problem size.
Table \ref{fig:bench} shows the timings of the various
solvers to solve different problem sizes.

@@

\section{Application: compact finite difference on multiple GPUs}

@@

\begin{equation}
    f^{\prime}_1 + 2f^{\prime}_2 = \frac{-5f_1 + 4f_2 + f_3}{dx}
\end{equation}

The right boundary ($f^{\prime}_{n}$) is treated by utilizing
the negative complex-conjugate of the Fourier image of the stencil
at $f^{\prime}_1$:

\begin{equation}
    f^{\prime}_{n} + 2f^{\prime}_{n-1s}
    =
    \frac{5f_{n} - 4f_{n-2} 1 f_{n-2}}{dx}
\end{equation}

This boundary treatment maintains the tridiagonal identity
of the syste to be solved, which then has the form:

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/computational-domain.eps}
\end{center}
\caption{Computational domain in 3-D}
\label{fig:computational-domain}
\end{figure}

@@

For multiple GPUs, we use the algorithm
proposed by Mattor et al. \cite{mattor1995algorithm}.
Here, we distribute the tridiagonal system among
$p$ procesors.
We solve for the ``local''  
portion of the global tridiagonal system
by first solving the following three ``local''
tridiagonal systems:

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
x_{r,1}^p \\
x_{r,2}^p \\
x_{r,3}^p \\
x_{r,4}^p \\
\vdots \\
x_{r,m}^p
\end{bmatrix}
=
\begin{bmatrix}
r_1^p \\
r_2^p \\
r_3^p \\
r_4^p \\
\vdots \\
r_m^p
\end{bmatrix} & \label{eqn:primary-system} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
u_1^p \\
u_2^p \\
u_3^p \\
u_4^p \\
\vdots \\
u_m^p
\end{bmatrix}
=
\begin{bmatrix}
-a_1^p \\
0 \\
0 \\
0 \\
\vdots \\
0
\end{bmatrix} & \label{eqn:secondary-system-1} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
l_1^p \\
l_2^p \\
l_3^p \\
l_4^p \\
\vdots \\
l_m^p
\end{bmatrix}
=
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
\vdots \\
-c_m^p
\end{bmatrix} & \label{eqn:secondary-system-2}
\end{align}

and additionally, the following ``reduced'' system:

@@

We call the system in Equation \ref{eqn:primary-system}
the ``primary'' system, and the systems in
Equations
\ref{eqn:secondary-system-1} and \ref{eqn:secondary-system-2}
the ``secondary'' systems.
The local part of the solution to the ``global'' tridiagonal system
is obtained as a linear combination of
the solutions to the primary and secondary sytems:

@@

\begin{enumerate}
    \item Computation of the RHS (requires communication)
    \item Solving the primary and secondary systems
    \item Constructing and solving the reduced systems,
        and scattering the paramters $\alpha^p$ and $\beta^p$ (requires communication)
    \item Summing the solutions (Equation \ref{eqn:sum-of-systems})
\end{enumerate}

Each of these is performed on the GPU,
one of the primary concerns is to eliminate
CPU-GPU transfers as much as possible in the solution process.

The right-hand side computation involves stencil computation,
which requires communication at the boundaries.
Communication of non-contiguous MPI
data types is expensive on the GPU,
so we have dedicated
contiguous halo arrays for performing swaps in each direction

@@

% Time for dfdx, dfdy, dfdz at various problem sizes
% and nprocs

% Breakdown - time allocated for each step

% Weak scaling

% Strong scaling

>>>>

\begin{abstract}
    In this paper, we consider the solution of linear
    systems of the form $A\bm{x} = \bm{d}$,
    where $A$ is \emph{near-Toeplitz} tridiagonal.
    These systems appear in numerical schemes
    such as
    Alternating Direction Implicit (ADI) methods,
    compact finite differences,
    and one-dimensional ordinary
    and partial differential equations.
    We develop a GPU solver that takes advantage
    of the known structure of the matrix
    to reduce the number of computations
    and improve memory access.
    Our solver is able to outperform
    the NVIDIA CUSPARSE and
    multi-core Intel MKL tridiagonal solvers
    (on up to 16 cores).
    We also describe the implementation of a multi-GPU solver
    using MPI and CUDA,
    and show scalability for up to 64 GPUs
    for a 3-D compact finite difference application.
\end{abstract}

@@

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}
%
where, in general
\begin{align*}
    & a_0 \neq a_n & \\
    & b_1 \neq b_0 \neq b_n &\\
    & c_1 \neq c_0 &
\end{align*}

@@

Such tridiagonal systems
appear in several numerical schemes
in Computational Fluid Dynamics (CFD)
such as
compact finite differences \cite{lele1992compact},
Alternating Direction Implicit (ADI) methods \cite{1955ADI}, and
numerical solutions of one-dimensional differential equations.
In such applications,
the evaluation of the tridiagonal systems constitutes
a significant portion of the runtime.
Recently, there has been considerable interest in the CFD community
in developing general tridiagonal solvers that
exploit the massive parallelism afforded by
Graphics Processing Units (GPUs)
to speed up the computations
\cite{tutkun2012gpu}
\cite{esfahanian2014efficient}
\cite{GoSt11CR}.
The performance of tridiagonal solvers on the GPU
has been shown to depend on several factors,
such as
choice of algorithm and the associated complexity,
thread utilization,
use of the available memory hierarchy,
and synchronization and control costs
\cite{Zhang2010FTS}.
Many approaches described in the literature
seek to make trade-offs between these factors
to arrive at a solver that best fits the requirements.
But to our knowledge,
the algorithms proposed so far
are applicable to general tridiagonal systems,
and make no attempt to take advantage of
the known structure of the tridiagonal matrix.
The contribution of this paper is twofold:
First, we describe the implementation of a GPU solver
for the specific case of near-Toeplitz systems
that uses the known structure to reduce
computations and memory accesses.
Secondly, we describe the implementation for
for multi-GPU clusters using MPI and CUDA,
and apply it to the evaluation of compact finite differences.

This paper is organized as follows:
Section \ref{sec:preliminaries}
describes algorithms for solving tridiagonal systems on GPUs.
The cyclic reduction algorithm in particular, is discussed in detail,
and issues with its implementation on GPUs is discussed.
Section \ref{sec:proposed-algorithm}
describes our proposed solver for near-Toeplitz tridiagonal systems.
Section \ref{sec:gpu-implementation}
discusses the implementation of the algorithm on GPUs.
Section \ref{sec:results-single-gpu}
provides an overview of the performance of our solver,
and compares it with other implementations.
Section \ref{sec-compact-finite-differences}
describes the implementation on multiple GPUs,
and application to the solution of compact finite differences.

@@

Cyclic reduction is applicable to the solution of
general linear systems of the form $A\bm{x} = \bm{d}$,
where $A$ is a tridiagonal matrix with diagonals
$\bm{a}$, $\bm{b}$ and $\bm{c}$
(Equation \ref{eqn:general-tridiagonal-system}).
The algorithm consists of two phases:
\emph{forward reduction} and \emph{backward substitution}
\ref{fig:cyclic-reduction}.
In the forward reduction phase,
every even-indexed equation $i$
is expressed as a
linear combination of equations $i$, $i-1$ and $i+1$,
yielding a new tridiagonal system of
$n/2$ equations in $n/2$ unknowns
(Equations \ref{eqn:forward-reduction-1} - \ref{eqn:forward-reduction-4}).
The process is repeated until a system of
2 equations in 2 unknowns is left.

\begin{align} 
& a^{\prime}_i = -a_{i-1}k_1 \
    \label{eqn:forward-reduction-1}& \\
& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 \
    \label{eqn:forward-reduction-2}& \\
& c^{\prime}_i = -c_{i+1}k_2 \
    \label{eqn:forward-reduction-3}& \\
& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
    \label{eqn:forward-reduction-4}&
\end{align}
%
where,

@@

\begin{align} \label{eqn:backward-substitution}
x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
    c^{\prime}_ix_{i+1}}{b^{\prime}_i}
\end{align}
%
For the last index $i=n$,
the forward reduction step is instead:
\begin{align} \label{eqn:forward-reduction-last}
    & a^{\prime}_n = -a_{n-1}k_1 & \\
    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
    & d^{\prime}_n = d_n - d_{n-1}k_1&
\end{align}
%
And for $i=1$, the backward substitution step is instead:
\begin{align} \label{eqn:backward-substitution-first}
x_1 = \frac{d^{\prime}_1 - c^{\prime}_1x_{2}}{b^{\prime}_1}
\end{align}

@@

Writing applications that effectively exploit the GPU
requires an understanding of the underlying architecture.
We provide the pertinent details here,
and refer to \cite{GPUcomputingera} for a more complete picture.
Our tests are performed on the NVIDIA Tesla K20 and K40 GPUs,
built on the ``Kepler'' architecture,
an overview of which is available at \cite{Keplerwhitepaper}.

@@

\begin{itemize}
    \item [] $a_2 = a_3 = a_4 = \hdots \equiv a_0$
    \item [] $b_2 = b_3 = b_4 = \hdots \equiv b_0$
    \item [] $c_2 = c_3 = c_4 = \hdots \equiv c_0$
\end{itemize}
%
we observe that the resulting coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
correspond to the coefficients of a tridiagonal matrix with
exactly the near-Toeplitz structure of $A$.
This form-preserving property of cyclic reduction
has been reported for block Toeplitz tridiagonal systems
by Bini et al.

@@

\begin{equation}
d^{\prime}_i = d_i - d_{i-1}k_1^{m}  - d_{i+1}k_2^{m}
\label{eqn:precomputed-forward-reduction-step}
\end{equation}
%
And the $m^{th}$ backward substitution step is:

\begin{equation}
x_i = \frac{d^{\prime}_i - a^mx_{i-1} - \
    c^{m}x_{i+1}}{b^m}
\label{eqn:precomputed-backward-substitution-step}
\end{equation}
%
where $a^m$, $b^m$, $c^m$, $k_1^m$ and $k_2^m$
are the appropriate values
from the precomputed coefficient arrays.
In practice,
the backward substitution step
can overwrite the right-hand side vector $\bm{d}$
with the solution values.

@@

Following the algorithm described in \ref{sec:proposed-algorithm},
the forward reduction coefficients
are first computed and loaded into global memory.
The right hand sides are stored in a single
contiguous region global memory.
In both approaches,
we map each of the $N_{rhs}$ tridiagonal systems
to a thread block,
and equations or \emph{indices} of a system
to individual threads within a block
(Figure \ref{fig:gpu-mapping}).

\begin{figure}[h!]
\begin{center}
\includegraphics[width=400pt]{img/global-and-shared.eps}
\end{center}
\caption{Shared memory (top) and Global memory (bottom) implementations}
\label{fig:global-and-shared}
\end{figure}

@@

This implementation works entirely on the GPU's global memory,
Here, we define two kernels - one for the forward reduction step,
and the other for the backward substitution step.
Each kernel is called $log_2(n)-1$ times.
An extra call to the
forward reduction kernel performs
the two-by-two solve.
At each step,
the size of the grid block is determined by the \emph{stride}
between elements accessed at that step.
For the forward reduction phase,
we use $n/2$ threads per block for the first step,
$n/4$ threads for the second step, and so on.
The pattern is reversed for the backward substitution phase,
beginning with 2 threads per block for the first step.
Although this ensures that there are no inactive threads
at any stage,
the occupancy of the GPU is still very low
during the end of forward reduction
and the end of backward substitution.

The precomputed coefficient arrays and right hand
are accessed by the kernels from global memory.
The kernel suffers from
strided memory access for the right-hand side,
but the precomputed coefficient values are accessed
without major coalescing problems.
Further, by precomputing the forward reduction coefficients,
we greatly reduce the number of computations
(and thus the number of uncoalesced memory accesses).

@@

In the shared memory approach (Figure \ref{fig:global-and-shared}),
we launch a single kernel to perform the entire
cyclic reduction solve.
The kernel is launched with $n/2$ threads per block.
Each thread block allocated a block of shared memory of size $n/2$.
Each thread performs the first reduction step
(Equation \ref{eqn:precomputed-forward-reduction-step})
by accessing the required values
$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
storing the result in shared memory.
In subsequent reduction steps,
$d_i$, $d_{i-1}$ and $d_{i+1}$
are accessed from shared memory,
avoiding the uncoalesced global memory accesses
seen in the global memory implementation.
In each back substitution step,
threads overwrite the existing values in shared memory
with the values of the solution.
In the final step,
shared memory is filled completeley
with the even-indexed solution values.
Each thread then computes an odd-indexed solution value,
storing it directly in global memory
and copies the even-indexed solution value
from shared memory to global memory.
The entire solution is done in a single kernel launch
to hide the latency of
memory transfers between global and shared memory.
Explicit synchronization between the threads of a block
is required within the kernel at the end of each step.

The shared memory implementation suffers from two major issues:
first, the number of active threads is halved at each forward reduction step,
(and subsequently doubled at each backward reduction step).
Synchronization between threads of a block is necessary at each step.
Thus, a significant portion of the kernel execution time is
spent by idle threads waiting for active threads to complete execution.
Secondly, the strided access to the
right-hand side values leads to \emph{bank-conflicts}.
However, the number of bank conflicts is significantly smaller
than in cyclic reduction implementations for general tridiagonal systems,
due the reduced number of computations performed.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
\caption{Comparison of global memory and shared
    memory implementations of NEATO (2D problems)}
\label{fig:global-vs-shared-2d}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=1.0\linewidth]{fig/global-vs-shared-3d.eps}
\caption{Comparison of global memory and shared memory
    implementations of NEATO (2D problems)}
\label{fig:global-vs-shared-3d}
\end{center}
\end{figure}

@@

In Figures \ref{fig:bench-2d} and \ref{fig:bench-3d},
we provide the relative performance of
Intel MKL and CUSPARSE solvers and compare against
the NEATO shared memory implementation.
The Intel MKL solver can use multiple threads
to solve several systems simultaneously.
The CUSPARSE solver uses a hybrid algorithm composed
of Cyclic Reduction and Parallel Cyclic Reduction (CR+PCR).
The relative performance for
each problem size is obtained by
normalizing the solver timings
by the timing for the fastest solver for that problem size.
Table \ref{table:bench} shows the timings of the various
solvers to solve different problem sizes.

@@

%----------------------------------------------------------------------
\section{Application: compact finite difference on multiple GPUs}
\label{sec-compact-finite-differences}
%----------------------------------------------------------------------

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/computational-domain.eps}
\caption{Computational domain in 3-D}
\label{fig:computational-domain}
\end{center}
\end{figure}

@@

\begin{equation}
    f^{\prime}_1 + 2f^{\prime}_2 = \frac{-5f_1 + 4f_2 + f_3}{dx}
\end{equation}
%
The right boundary ($f^{\prime}_{n}$) is treated by utilizing
the negative complex-conjugate of the Fourier image of the stencil
at $f^{\prime}_1$:

\begin{equation}
    f^{\prime}_{n} + 2f^{\prime}_{n-1s}
    =
    \frac{5f_{n} - 4f_{n-2} 1 f_{n-2}}{dx}
\end{equation}
%
This boundary treatment maintains the tridiagonal identity
of the system to be solved, which then has the form:

@@

@@

For multiple GPUs, we use the algorithm
proposed by Mattor et al. \cite{mattor1995algorithm}.
Here, we distribute the tridiagonal system among
$p$ processors.
We solve for the ``local''  
portion of the global tridiagonal system
by first solving the following three ``local''
tridiagonal systems:

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
x_{r,1}^p \\
x_{r,2}^p \\
x_{r,3}^p \\
x_{r,4}^p \\
\vdots \\
x_{r,m}^p
\end{bmatrix}
=
\begin{bmatrix}
r_1^p \\
r_2^p \\
r_3^p \\
r_4^p \\
\vdots \\
r_m^p
\end{bmatrix} & \label{eqn:primary-system} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
u_1^p \\
u_2^p \\
u_3^p \\
u_4^p \\
\vdots \\
u_m^p
\end{bmatrix}
=
\begin{bmatrix}
-a_1^p \\
0 \\
0 \\
0 \\
\vdots \\
0
\end{bmatrix} & \label{eqn:secondary-system-1} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
l_1^p \\
l_2^p \\
l_3^p \\
l_4^p \\
\vdots \\
l_m^p
\end{bmatrix}
=
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
\vdots \\
-c_m^p
\end{bmatrix} & \label{eqn:secondary-system-2}
\end{align}
%
and additionally, the following ``reduced'' system:

@@

We call the system in Equation \ref{eqn:primary-system}
the ``primary'' system, and the systems in
Equations
\ref{eqn:secondary-system-1} and \ref{eqn:secondary-system-2}
the ``secondary'' systems.
The local part of the solution to the ``global'' tridiagonal system
is obtained as a linear combination of
the solutions to the primary and secondary systems:

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=200pt]{img/constructing-reduced-system.eps}
\caption{Construction of the reduced system and scattering of parameters}
\label{fig:reduced-system}
\end{center}
\end{figure}

In this section we describe the implementation of
a CUDA+MPI solver for compact finite differences.
The algorithm 
consists of the following sequence of steps:

\begin{enumerate}
\item Computation of the RHS (requires communication)
\item Solving the primary and secondary systems
\item Constructing and solving the reduced systems,
    and scattering the parameters $\alpha^p$ and $\beta^p$ (requires communication)
\item Summing the above solutions (Equation \ref{eqn:sum-of-systems})
\end{enumerate}

Each of these is performed on the GPU,
and one of the primary concerns is to eliminate
CPU-GPU transfers as much as possible in the solution process.

The right-hand side evaluation is a pointwise stencil computation,
which requires communication of the function values at the boundaries.
This is achieved using a halo-swapping technique
with dedicated contiguous halo arrays for each subdomain,
as communication of non-contiguous MPI
data types is expensive on the GPU.
After the halo information is exchanged,
a pointwise stencil kernel computes the right-hand side values.

The secondary systems are set up and solved only once
as it is dependent only on the
coefficients of the tridiagonal matrix.
As the computation is offline,
any convenient method can be used.
The primary system is solved at every time step
for different right hand sides,
and is evaluated using the NEATO solver.

The set up and solution of the reduced system
requires global communication of the boundary information,
which may be associated with a high overhead.
While several techniques have been proposed to implement
this type of communication pattern efficiently,
we find it sufficient for moderately large problems
to communicate all information to rank 0,
using \texttt{MPI\_Gather}, and distribute the results
back using \texttt{MPI\_Scatter} (Figure \ref{fig:reduced-system}).

Finally, the summing of the solutions
is a pointwise operation that is easily implemented
on the GPU.

For evaluation of compact finite differences in different
coordinate directions,
a local permutation of the data is performed.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-256.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $256^3$}
\label{fig:strong-scaling-256}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/strong-scaling-512.eps}
\caption{Strong scaling for multi-GPU compact finite difference, problem size: $512^3$}
\label{fig:strong-scaling-512}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-128.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $128^3$}
\label{fig:weak-scaling-128}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/weak-scaling-256.eps}
\caption{Weak scaling for multi-GPU compact finite difference, problem size: $256^3$}
\label{fig:weak-scaling-256}
\end{center}
\end{figure}
